Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2021-02-11 manuscript_Huang_rev.tex
files updated.

In addition to the proper normalization, we need to consider self-loops that may affect the numerical estimation of $\mathbf{W}$ especially for time series from continuous systems with a given sampling rate. For example, it has been demonstrated in Ref.~\cite{zhangSciRep2017} that there are about $99\%$ self-loops in time series of the R\"ossler system when integrated with a step size $h = 0.01$, while the about $1\%$ non-self-loops are hidden in $\mathbf{W}$. From this example, one may easily see that self-loops are related to the serial correlation of the time series under study \cite{BorgesAMC2019}. {\color{red} From the viewpoint of statistical measures of complex network theory}, neglecting self-loops in $\mathbf{W}$ emphasizes the transition behavior between different patterns, which has therefore been adopted in most existing studies of complex networks for computational simplicity and theoretical concerns \cite{CostaADPhy2007}. Accordingly, we also remove self-loops in the present study before computing $\mathbf{W}$ while acknowledging that in other cases of {\color{red} highly correlated} stochastic processes and discrete systems, self-loops have been included in the analysis showing some interesting results \cite{BorgesAMC2019}. Arguably, whether to remove or to consider self-loops depends on the particular process under study. In any of the applications discussed in the rest of this paper, our results do not change qualitatively when including self-loops {\color{red}(see Supplementary Material Sec.~SM1 with Figs.~S1 to S3 for some illustrative examples).}

@@

The situation becomes even more challenging when our modified SCMs are considered, since they are based on transition frequencies instead of pattern frequencies. For $D!$ different ordinal patterns for embedding dimension $D$, there may exist $D!^2$ possible source--destination pairs. However, if we consider $\tau=1$ and transitions corresponding to shifting the time index by one time step only, it is clear that from the $D$ components of the source permutation, $D-1$ also contribute to the corresponding destination pattern without changing their mutual order relationships. Hence, the new state contributing to the destination pattern can only lead to at most $D$ different ordinal patterns, i.e., we have at most $D\cdot D!$ different pattern transitions. However, for other cases like $\tau>1$ and/or larger time shifts between ordinal patterns to be considered, up to $D!^2$ different pairs of patterns and, hence, transitions could occur in the worst case. In general, we conclude that the far higher number of pattern transitions as compared to patterns themselves puts severe restrictions to the choice of $D$ in practical applications of the new SCMs in the context of time series of a given length $L$.
}

@@

\textcolor{red}{More importantly, for sufficiently large $D$ the resulting complexity--entropy pairs} essentially cover exclusively the left uprising branch of the curve for the SCM \textcolor{red}{accounting for} transition frequencies (Figs.~\ref{fig:CElogistic}(b, e) and \ref{fig:CElogistic}(c, f)), which distinctively differs from the original permutation entropy based SCM. \textcolor{red}{In this regard, one may note that this branch implies an almost linear dependency between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig.~\ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig.~\ref{fig:CElogistic}(f)), which especially applies for small entropy values within an increasingly wide range as $D$ increases. This linear increase of the maximum complexity versus entropy curve for sufficiently large embedding dimension can be expected from theoretical considerations~\cite{SMCbook2010}.}

@@

The first example originates from a laboratory fluid experiment of baroclinic instability which is used to study patterns of cyclones and anticyclones in the Earth's atmosphere~\cite{Read_jfm_1992,ZouEPJST2008}. Depending on the parameters rotation rate, temperature difference, viscosity and fluid density, this experimental system exhibits a rich variety of flow regimes. In particular, we focus on the following regimes: 
\begin{enumerate}[(i)]
\item {Stable fluid (SW):} The temperature signal of the stable flow exhibits periodic oscillations as the wave drift patterns arrive at the fixed point of measurements. 
\item {Quasi-periodic 2-frequency amplitude vacillation (AV-2):} This case is identified as a 2-frequency quasi-periodic amplitude flux. The wave drift is composed of slow and regular oscillations in the temperature signal, but a fast modulation in the amplitude is also visible. AV-2 is characterized by a periodic growth and drop in wave amplitude with little change in waveform. 
\item {Quasi-periodic 3-frequency amplitude vacillation (AV-3):} For the sake of completeness, we also analyze a quasi-periodic 3-frequency amplitude vacillation time series. Unlike for the other studied cases, this time series has not been obtained from experimental results, but from a three-dimensional direct numerical simulation of the air-filled rotating baroclinic instability experiment~\cite{Read_jfm_1992}. 
\item {Modulated amplitude vacillation (MAV):} This case is identified as a low-dimensional stream, chaotically modulated and with waves of varying amplitudes. Amplitude modulation results in a complex temperature dynamics.
\end{enumerate}
{\color{red}One temperature time series of each flow regime has been measured at $2$~s resolution for periods of up to $9 \cdot 10^4$s. Therefore, each recording consists of $N = 45,000$ data points. Example segments of the resulting time series are shown in Fig.~\ref{tseriesFluid} for illustrative purposes. Further details on the lower-dimensional chaotic properties have been reported in Refs.~\cite{Read_jfm_1992,thiel2004a}. }
\begin{figure}
	\centering 
	\includegraphics[width=\columnwidth]{TS_fluid.pdf}
\caption{\small{Sample segments of the temperature recordings from the flow experiment: (a) SW, (b) AV-2, (c) AV-3 and (d) MAV. The vertical axis i\textcolor{red}{describes the} temperature $T$. } \label{tseriesFluid}}
\end{figure}

>>>>

In addition to the proper normalization, we need to consider self-loops that may affect the numerical estimation of $\mathbf{W}$ especially for time series from continuous systems with a given sampling rate. For example, it has been demonstrated in Ref.~\cite{zhangSciRep2017} that there are about $99\%$ self-loops in time series of the R\"ossler system when integrated with a step size $h = 0.01$, while the about $1\%$ non-self-loops are hidden in $\mathbf{W}$. From this example, one may easily see that self-loops are related to the serial correlation of the time series under study \cite{BorgesAMC2019}. {\color{red} From the viewpoint of statistical measures of complex network theory}, neglecting self-loops in $\mathbf{W}$ emphasizes the transition behavior between different patterns, which has therefore been adopted in most existing studies of complex networks for computational simplicity and theoretical concerns \cite{CostaADPhy2007}. Accordingly, we also remove self-loops in the present study before computing $\mathbf{W}$ while acknowledging that in other cases of {\color{red} highly correlated} stochastic processes and discrete systems, self-loops have been included in the analysis showing some interesting results \cite{BorgesAMC2019,PessaPRE2019}. Arguably, whether to remove or to consider self-loops depends on the particular process under study. In any of the applications discussed in the rest of this paper, our results do not change qualitatively when including self-loops {\color{red}(see Supplementary Material Sec.~SM1 with Figs.~S1 to S3 for some illustrative examples).}

@@

The situation becomes even more challenging when our modified SCMs are considered, since they are based on transition frequencies instead of pattern frequencies. For $D!$ different ordinal patterns for embedding dimension $D$, there may exist $D!^2$ possible source--destination pairs. However, if we consider $\tau=1$ and transitions corresponding to shifting the time index by one time step only, it is clear that from the $D$ components of the source permutation, $D-1$ also contribute to the corresponding destination pattern without changing their mutual order relationships \cite{BorgesAMC2019,PessaPRE2019}. Hence, the new state contributing to the destination pattern can only lead to at most $D$ different ordinal patterns, i.e., we have at most $D\cdot D!$ different pattern transitions. However, for other cases like $\tau>1$ and/or larger time shifts between ordinal patterns to be considered, up to $D!^2$ different pairs of patterns and, hence, transitions could occur in the worst case. In general, we conclude that the far higher number of pattern transitions as compared to patterns themselves puts severe restrictions to the choice of $D$ in practical applications of the new SCMs in the context of time series of a given length $L$.
}

@@

\textcolor{red}{More importantly, for sufficiently large $D$ the resulting complexity--entropy pairs} essentially cover exclusively the left uprising branch of the curve for the SCM \textcolor{red}{accounting for} transition frequencies (Figs.~\ref{fig:CElogistic}(b, e) and \ref{fig:CElogistic}(c, f)), which distinctively differs from the original permutation entropy based SCM. \textcolor{red}{In this regard, one may note that this branch implies an almost linear dependency between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig.~\ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig.~\ref{fig:CElogistic}(f)), which especially applies for small entropy values within an increasingly wide range as $D$ increases. This linear increase of the maximum complexity versus entropy curve for sufficiently large embedding dimension is expected from theoretical considerations~\cite{SMCbook2010}.}

@@

The first example originates from a laboratory fluid experiment of baroclinic instability which is used to study patterns of cyclones and anticyclones in the Earth's atmosphere~\cite{Read_jfm_1992,ZouEPJST2008}. Depending on the parameters rotation rate, temperature difference, viscosity and fluid density, this experimental system exhibits a rich variety of flow regimes. In particular, we focus on the following regimes: 
\begin{enumerate}[(i)]
\item {Stable fluid (SW):} The temperature signal of the stable flow exhibits periodic oscillations as the wave drift patterns arrive at the fixed point of measurements. 
\item {Quasi-periodic 2-frequency amplitude vacillation (AV-2):} This case is identified as a 2-frequency quasi-periodic amplitude flux. The wave drift is composed of slow and regular oscillations in the temperature signal, but a fast modulation in the amplitude is also visible. AV-2 is characterized by a periodic growth and drop in wave amplitude with little change in waveform. 
\item {Quasi-periodic 3-frequency amplitude vacillation (AV-3):} For the sake of completeness, we also analyze a quasi-periodic 3-frequency amplitude vacillation time series. Unlike for the other studied cases, this time series has not been obtained from experimental results, but from a three-dimensional direct numerical simulation of the air-filled rotating baroclinic instability experiment~\cite{Read_jfm_1992}. 
\item {Modulated amplitude vacillation (MAV):} This case is identified as a low-dimensional stream, chaotically modulated and with waves of varying amplitudes. Amplitude modulation results in a complex temperature dynamics.
\end{enumerate}
{\color{red}One temperature time series of each flow regime has been measured at $2$~s resolution for periods of up to $9 \cdot 10^4$s. Therefore, each recording consists of $N = 45,000$ data points. Example segments of the resulting time series are shown in Fig.~\ref{tseriesFluid} for illustrative purposes. Further details on the lower-dimensional chaotic properties have been reported in Refs.~\cite{Read_jfm_1992,thiel2004a}. }
\begin{figure}
	\centering 
	\includegraphics[width=\columnwidth]{TS_fluid.pdf}
\caption{\small{Sample segments of the temperature recordings from the flow experiment: (a) SW, (b) AV-2, (c) AV-3 and (d) MAV. The vertical axis \textcolor{red}{describes the} temperature $T$. } \label{tseriesFluid}}
\end{figure}
